Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-01 doc/types-pldi2013/langdef.tex
updates

\end{tikzpicture}
}
\vspace*{-0.5cm}
\texcomment{
\framebox[7in]{
\placeat{2}{3}{bar}%
\placeat{3}{1}{3,1}%
\placeat{-3}{1}{-3,1} %
\makebox[0pt]{%
\hspace*{2in} \raisebox{3in}{---}%
} %
\makebox[0pt]{%
\hspace*{1in} \raisebox{1in}{foo}%
} %
\makebox[0pt]{%
\hspace*{2in} \raisebox{3in}{bar}%
}%
\rule{0.1pt}{6in}
}}
}
\caption{Type System and Operational Semantics Rules}
\label{fig:langrules}
\end{figure*}

@@

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  Types
include booleans, integers, tuples, and pointers.  Pointers
are annotated with a list of regions---a non-null pointer must point to a 
location that is contained in at least one of the regions. There is a special
type for {\em colorings}, which
are used to specify partitions of regions into subregions.
Functions in Legion are named, accept one or more arguments and
return a value of some type.  A function also declares the region
access privileges and coherence it requires.  A function may be executed
within the current task or run in parallel as a separate subtask---this decision is made
by the Legion task scheduler.  Function types are
universally quantified over all region names appearing in the type.
The final type is a region relationship, an instance of which captures a value and
one or more regions satisfying associated constraints.  A region relation instance
can be written into the heap and later read and
unpacked, giving new local names to the regions contained in
the instance.

To achieve high performance in the presence of aliasing, the Legion type system has 
been designed to efficiently leverage both static and dynamic checks.
Consequently, both the type system and the operational
semantics have been extended in interesting ways from a basic expression language.  We
describe these extensions in the next two subsections, and explain how they 
were chosen to preserve a straightforward proof of soundness for the type system.

@@

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$

In addition to the traditional type environment $\Gamma$, a Legion type judgment includes the
set of access privileges $\Phi$ for the logical regions in the expression as well as a set of
constraints $\Omega$ that must hold between those logical regions.

Both $\Phi$ and $\Omega$ are used in the heap access expressions: {\em read}, {\em write}, and {\em reduce}.
For a heap access to be valid, the the appropriate permission must exist for logical region(s) in the
pointer's type.  As regions are hierarchical, the exact region need not be named in $\Phi$ if permissions
exist for a logical region that is known to contain the pointer's region.  To simplify this check, we
define closure operations in Figure~\ref{fig:closure} that expand $\Omega$ into an $\Omega^*$, which is
then used to expand $\Phi$ into a $\Phi^*$ that does explicitly name every logical region for which
privileges are known to exist.

Core Legion provides no automatic pointer type conversions.  Valid pointers are created via the {\em new}
expression in specified region and invalid (i.e. {\em null}) pointers must also be typed.  Pointers can be
``upcast'' via the {\em upregion} expression, which uses the expanded constraints $\Omega^*$ to verify that
every possible region a pointer might point into is covered by at least one of the regions in the desired
pointer type.  In contrast, a ``downcast'' using the {\em downregion} expression must perform a dynamic
test to verify which region a run-time pointer value points into.  A static test similar to the one used
for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
have left it out for simplicity.

The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
pointers (which must point into the region being colored) to integer {\em color} values.  (For simplicity, 
Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
expression introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
subregion is disjoint from the other subregions and all are included in the original region.

The {\em pack} and {\em unpack} expressions allow the storage of pointers in the heap.  Logical region names
are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  A region relationship
is used instead to capture the (dynamic) physical regions while preserving the statically known constraints
between those regions.  Because the {\em pack} expression is able to verify these constraints hold when the 
region relationship value is created, the {\em unpack} expression is able to reintroduce these constraints on
the (fresh) logical region names given to the contents of an unwrapped region relationship value.

Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
verifies the functional requirement that the privileges held by a called task are always a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enable the hierarchical scheduling model used by the Legion runtime.

@@

Core Legion's operational semantics rules have the following form:
$$\opsenv{e}{v}{E}$$
and specify that, given a certain environment, the evaluation of expression $e$ yields a value $v$ and a
{\em dynamic memory trace} $E$ which is a sequence of all heap operations (reads, writes, reductions) performed
during the evaluation, including their location and the value read or written.

In addition to the usual value mappings for local variables $L$ and store (i.e. heap) locations $S$, the Legion
operational semantics environments includes an immutable heap typing $H$ that maps locations to types, and a region
mapping $M$ that maps logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
locations.

The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical fashion.  We therefore require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence with aliasing), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global scheduler is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints on the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em coherence} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

The complete operational semantics rules can be found in \cite{LegionTypes12}; we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
will be discussed in Section~\ref{sec:coherence}.)

The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  The new
expression is constrained to select a location (with the right heap typing) from the set that the specified logical
region has been mapped to.  (Although one would expect the new expression to also return a location that is not
currently in use, our soundness result does not require this additional constraint, so we have omitted it for
simplicity.)  Similarly, the downregion expression checks whether a location is within the set of locations the
logical region(s) have been mapped to, returning {\em null} if not.  As discussed above, the corresponding 
{\em upregion} expression checks are performed statically.

The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
existing coloring and returns one in which the specified location is given the specified color, replacing an
existing coloring for that location (if present).  The behavior of the newcolor expression is more subtle.
Since Legion supports the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
guaranteeing that any explicit colorings performed by the program are honored.

The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  Packing a region relationship
is just a matter of capturing the physical regions to which the logical regions are currently mapped, whereas 
unpacking takes those physical regions and gives them fresh logical names within the evaluation of the body.

Finally, although the semantics for calls to subtasks appear complicated at first glance, the bulk of the rules
are just handling the evaluation of each actual argument expression and the creation of the called subtask's 
environment (with potentially different names for logical regions).  The key difference in Legion task calls is the
handling of relaxed coherence.  Any location that falls within
a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the clobber set
$C'$, allowing non-deterministic evaluation within the subtask with respect to those locations.  All actual accesses
to those locations within the resulting memory trace ($E_{n+1}$) are marked with the corresponding coherence mode,
allowing the {\em valid\_interleave} check within the caller's environment to verify the consistency of the overall
memory trace.

@@

A key property of the Legion type system is that any expression that is well-typed is
guaranteed to access the heap only in ways permitted by the privileges under which it was typed.
A judgment $\privconsist{E}{\Phi}$ holds if all the memory operations in memory trace $E$ are of the types and
to locations covered by the privileges in $\Phi$:
\[ 
\begin{array}{l}
\privconsist{E}{\Phi} \Leftrightarrow \forall \epsilon \in E. \\
\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
\end{array}
\]

Our goal is to show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static privileges $\Phi$.
As usual, even stating such a claim requires that corresponding parts of the initial static and dynamic environments be consistent.
For our results, three consistency properties are needed:
\begin{itemize}
\item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$ under
which an expression was typed
\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment under which the expression was typed
\item store consistency, written $\storeconsist{S}{H}$, guarantees every location in the store $S$ contains a value consistent with the location's type in the heap typing $H$
\end{itemize}

\noindent Two additional properties are proven for each subexpression:
\begin{itemize}
\item result value consistency, written $\valueconsist{v}{T}$, guarantees any evaluation of an expression yields a value of the right type
\item memory trace consistency, written $\traceconsist{E}$, guarantees that all writes and reductions use values of the right types
\end{itemize}
Figure~\ref{fig:constprop} gives the precise definition of these properties.

@@

The proof, which can be found in \cite{LegionTypes12}, is lengthy but
straightforward due to the way in which the operational semantics were
constructed.  It proceeds by standard induction on the structure of the
derivation in the operational semantics, at each step showing that
subexpressions have consistent starting environments and that the
value and trace consistency of the subexpressions is sufficient to prove
the consistency of the parent expression.

>>>>

\end{tikzpicture}
}
\vspace*{-0.5cm}
\texcomment{
\framebox[7in]{
\placeat{2}{3}{bar}%
\placeat{3}{1}{3,1}%
\placeat{-3}{1}{-3,1} %
\makebox[0pt]{%
\hspace*{2in} \raisebox{3in}{---}%
} %
\makebox[0pt]{%
\hspace*{1in} \raisebox{1in}{foo}%
} %
\makebox[0pt]{%
\hspace*{2in} \raisebox{3in}{bar}%
}%
\rule{0.1pt}{6in}
}}
}
\caption{Type System and Operational Semantics}
\label{fig:langrules}
\end{figure*}

@@

Figure~\ref{fig:langdef} defines Core Legion, a subset of the full Legion language
that still illustrates all of the important issues.  
%Types include booleans, integers, tuples, and pointers.  
Pointers are annotated with a list of regions---a non-null pointer must point to a 
location that is contained in at least one of the regions. There is a special
{\em coloring} type, used to specify partitions of regions into subregions.
%Functions in Legion are named, accept one or more arguments and
%return a value of some type.  
Functions declare the region
access privileges and coherence they require.  Some function instances are {\em tasks}; a function may be executed
within the current task or run in parallel as a separate (sub)task---this decision is made
by the Legion task scheduler.  Function types are
universally quantified over all region names appearing in the type.
The other important type is a {\em region relationship}, an instance of which captures a value and
one or more regions satisfying associated constraints.  A region relation instance
can be written into the heap and later read and
unpacked, giving new local names to the regions contained in
the instance.

To achieve high performance in the presence of aliasing, the Legion type system
leverages both static and dynamic checks, resulting in interesting extensions beyond
a basic expression language to both the type system and the operational semantics.
The following two subsections illustrate these decisions and how they have been
chosen to preserve a straightforward proof of type soundness.

@@

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$
Besides a type environment $\Gamma$, type judgments include the
access privileges $\Phi$ for the logical regions in the expression $e$ as well as 
constraints $\Omega$ that must hold between those logical regions.

A representative selection of the type rules is given in Figure~\ref{fig:langrules}.  Both $\Phi$ and $\Omega$ are used in the heap access expressions {\em read}, {\em write}, and {\em reduce}.
A valid heap access has the needed permission for logical region(s) in the
pointer's type.  Note the exact region need not be named in $\Phi$ if permissions
exist for a logical region that provably contains the pointer's region.  To simplify this check, Figure~\ref{fig:closure}
defines closure operations $\Omega^*$ (all constraints implied by $\Omega$) and
$\Phi^*$ (all privileges implied by $\Phi$ and $\Omega^*$).

Core Legion provides no automatic pointer type conversions.  Valid pointers are created via the {\em new}
expression in the specified region.
% and invalid (i.e. {\em null}) pointers must also be typed.  
Pointers can be
``upcast'' via the {\em upregion} expression; the type rule verifies 
every possible region the pointer might point into is covered by at least one of the regions in the target
pointer type.  A ``downcast'' using the {\em downregion} expression performs a dynamic
test to verify the region a run-time pointer value points into.  
%A static test similar to the one used
%for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
%have left it out for simplicity.

Region constraints are introduced by the {\em partition} expression,
which makes use of {\em colorings}, a new type specific to Legion.  A coloring is an opaque mapping from
pointers (which must point into the region being colored) to integer {\em color} values.  
%(For simplicity, 
%Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
%an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
%coloring allows for more efficient generation (and storage) of colorings as needed.)  
Partitioning introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
subregion is disjoint from the others and all are subregions of the original region.

The {\em pack} and {\em unpack} expressions allow regions to be stored in the heap.
%  Logical region names are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  
A {\em region relationship} captures statically known constraints between regions;
because the {\em pack} rule verifies these constraints hold when the 
region relationship value is created,{\em unpack} can reintroduce these constraints on
the (fresh) logical region names when the value is read and unpacked from the heap.
Crucially for soundness,  region relationships do not capture privileges.
%
% FIXME  It would be nice to add some comments about unpack here, and how the constraints introduced by unpack
% must be sufficient to allow access to the new regions if they are to be used, which generally requires the type
% have a free region variable capturing subregioning relationships between the packed regions and some existing region
% in scope for which privileges are held.
% 
Finally, task calls verify the requirement that the privileges held by a called task are a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enabling hierarchical and distributed scheduling of tasks.

@@

Core Legion's operational semantics rules have the form
$$\opsenv{e}{v}{E}$$
and specify the evaluation of expression $e$ yields a value $v$ and a
{\em dynamic memory trace} $E$, which is a sequence of all heap operations (reads, writes, reductions) $e$ performs.
%during the evaluation, including their location and the value read or written.
In addition to  mappings for local variables $L$ and the store (i.e. heap) $S$, the 
environment includes an immutable heap typing $H$ assigning types to heap locations, and a region
mapping $M$ mapping logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
locations.

The environment also has a {\em clobber set} $C$.
We will show that parallel Legion programs can be scheduled in
a hierarchical fashion.  We therefore require an
operational semantics that is parallel and hierarchical.  Now
Legion allows limited nondeterminism, specifically because of relaxed coherence on
regions shared between parallel tasks and choices of reduction orderings.
Thus, heap accesses performed by one expression may observe side-effects from 
concurrent evaluations.  The traditional solution to this issue
is to use a top-down, small-step semantics with a global view of all concurrently executing expressions (e.g., \cite{CML}).  
%Rules are specified for which
%expression(s) are valid choices for the next small step of execution.
This approach is incompatible with a hierarchical view in which local
decisions are made with limited information.

Instead we use a bottom-up approach with big-step semantics.
Each expression has a clobber set $C$ of memory locations that might 
be modified by concurrent evaluations.  While a read from a location not in $C$ is {\em coherent}
(returning the most recently written value), the value read from a location $l$ in $C$ may be any
value.  Determing which value is deferred until all potentially-concurrent
modifications to $l$ are observable in the evaluation of a parent expression whose clobber set does
not include $l$ and a consistent interleaving of the subexpressions' memory traces is determined.
%With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
%made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
%used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

%FIXME  need to add much earlier that the complete report has all missing type and operational rules and proofs

We show the most interesting operational rules for Core Legion in
Figure~\ref{fig:langrules}.  The rules for the {\em read}, {\em
write}, and {\em reduce} show how dynamic memory traces are
constructed and how the result of a read expression is relaxed when
the location falls within the clobber set.  Additionally, the write
and reduce operations, with their two subexpressions, demonstrate how
the traces of subexpressions are interleaved within the environment
(and clobber set) of the parent expression.  (A formal definition of
the {\em valid\_interleave} predicate can be found in
Figure~\ref{fig:validinterleave}, and will be discussed in
Section~\ref{sec:coherence}.)

%The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  
The {\em new} expression selects a location (with the right heap typing) from those assigned to 
the logical region argument.
%(Although one would expect the new expression to also return a location that is not
%currently in use, our soundness result does not require this additional constraint, so we have omitted it for
%simplicity.)  
Similarly, {\em downregion} checks whether a location is within the set assigned to the logical region.
%logical region(s) have been mapped to, returning {\em null} if not.  
As discussed above, 
{\em upregion} checks are performed statically.

% FIXME The explanation of newcolor is confusing.
%
%The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
%operational semantics rules for the {\em newcolor} and {\em color} expressions.  
The {\em color} expression updates an existing coloring with a new color for a location, replacing any
existing coloring for that location.  The behavior of {\em newcolor} is subtler.
Legion supports first partitioning regions and then allocating data in the subregions.
%the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
%on subregions, 
Thus, each subregion resulting from a partitioning may need an arbitrary number of locations that are
unknown at partitioning time.  The semantics of newcolor allows this while
guaranteeing all explicit colorings performed by the program are honored.

%The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  
Packing a region relationship captures the physical regions to which the logical regions are mapped;
unpacking simply gives those physical regions fresh logical names.

Finally, calls to subtasks evaluate actual arguments and create the subtasks's
environment (with potentially different names for logical regions).  A key point in Legion task calls is the
handling of relaxed coherence.  Any location that falls within
a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the clobber set
$C'$, allowing non-deterministic evaluation within the subtask with respect to those locations.  All actual accesses
to those locations within the resulting memory trace ($E_{n+1}$) are marked with the corresponding coherence mode,
allowing the {\em valid\_interleave} check within the caller's environment to verify the consistency of the overall
memory trace.

@@

Our first results shows a well-typed expression accesses the heap in ways consistent with its static privileges.
A judgment $\privconsist{E}{\Phi}$ holds if memory operations in memory trace $E$ have types and
locations covered by privileges $\Phi$:
\[ 
\begin{array}{l}
\privconsist{E}{\Phi} \Leftrightarrow \forall \epsilon \in E. \\
\quad\quad (\epsilon = read(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reads(r) \in \Phi)\ \wedge \\
\quad\quad (\epsilon = write(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge writes(r) \in \Phi)\ \wedge \\
\quad\quad (\epsilon = reduce_{id}(l, c, v) \Rightarrow \exists r. l \in M(r) \wedge reduces_{id}(r) \in \Phi)
\end{array}
\]

\noindent
%We must show $\privconsist{E}{\Phi}$ holds for any execution of an expression $e$ with static privileges $\Phi$.
As usual, the soundness claim is proven assuming the initial type and execution environments are consistent.
For our results, three consistency properties are needed:
\begin{itemize}
\item mapping consistency, written $\mapconsist{\Omega}$, guarantees a region mapping $M$ satisfies the region constraints $\Omega$% under which an expression was typed
\item local value consistency, written $\localconsist{L}{\Gamma}$, guarantees local values in $L$ have types consistent with the environment $\Gamma$
%under which the expression was typed
\item store consistency, written $\storeconsist{S}{H}$, guarantees locations in $S$ have values consistent with heap typing $H$
\end{itemize}

\noindent Two additional properties are proven for each subexpression:
\begin{itemize}
\item result value consistency, written $\valueconsist{v}{T}$, guarantees any evaluation of an expression yields a value of the right type
\item memory trace consistency, written $\traceconsist{E}$, guarantees that all writes and reductions use values of the right types
\end{itemize}
Figure~\ref{fig:constprop} defines these properties.

@@

The proof, which can be found in the submitted appendix, is lengthy but
straightforward due to the compositional operational semantics.
%It proceeds by standard induction on the structure of the
%derivation in the operational semantics, at each step showing that
%subexpressions have consistent starting environments and that the
%value and trace consistency of the subexpressions is sufficient to prove
%the consistency of the parent expression.
